---
title: "DATA 589 Project"
author: "Christopher Mulya | Natalie Coutinho | Pranav Kumar Mahesh"
date: "2024-04-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load required packages
suppressWarnings({
  library(sp)
  library(sf)
  library(spatstat)
  library(maptools)
})

# Import the BC Covariates dataset
suppressWarnings({
  load("BC_Covariates.Rda")
})

library(readxl)

# Read the Excel file
suppressWarnings({
  df <- read_excel("Moose_data.xlsx")
})

# Extract decimalLatitude and decimalLongitude columns
latitude_column <- df$decimalLatitude
longitude_column <- df$decimalLongitude
province <- df$stateProvince

# Create a new data frame with latitude and longitude columns
coordinates_df <- data.frame(province = province, latitude = latitude_column, longitude = longitude_column)

# Load the required library
suppressWarnings({
  library(dplyr)
})

# Filter coordinates_df to only include observations from British Columbia (BC)
coordinates <- coordinates_df %>%
  filter(province == "British Columbia" | province == "Bc" | province == "Canada - British Columbia (BC)")


coordinates <- na.omit(coordinates)

# Print the filtered data frame for verification
head(coordinates)
```

```{r}
# Identify the class of each element within the list
sapply(DATA, class)
```

```{r}
# Extracting each element from DATA
Forest <- DATA$Forest
Window <- DATA$Window
Elevation <- DATA$Elevation
HFI <- DATA$HFI
Dist_Water <- DATA$Dist_Water
```

```{r}
# Plot latitude and longitude coordinates
plot(coordinates$longitude, coordinates$latitude, 
     xlab = "Longitude", ylab = "Latitude",
     main = "Scatter Plot of Latitude and Longitude")
```

```{r}
# Create a SpatialPoints object from latitude and longitude coordinates
coordinates_sp <- SpatialPoints(coords = coordinates[, c("longitude", "latitude")], 
                                proj4string = CRS("+proj=longlat +datum=WGS84"))

# Define the BC Albers projection
bc_albers_proj <- CRS("+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000 +y_0=0 +datum=NAD83 +units=m +no_defs")

# Transform the coordinates to BC Albers projection
coordinates_bc_albers <- spTransform(coordinates_sp, bc_albers_proj)

# Plot the transformed coordinates
plot(coordinates_bc_albers, 
     xlab = "BC Albers Easting (m)", ylab = "BC Albers Northing (m)",
     main = "BC Albers - Projection of Latitude and Longitude Coordinates")
```

```{r}
# Extract coordinates from the spatial object
coords <- coordinates(coordinates_bc_albers)

# Convert the coordinates to a dataframe
coordinates_df <- as.data.frame(coords)

# Remove duplicate points from the dataframe
coordinates_unique <- unique(coordinates_df)

# Extract x and y coordinates
x_coords <- coordinates_unique$coords.x1
y_coords <- coordinates_unique$coords.x2

# Create the window object
win <- as.owin(Window)

# Check which points are inside the window
inside_window <- inside.owin(x_coords, y_coords, win)

# Subset the coordinates dataframe to keep only the points inside the window
coordinates_inside_window <- coordinates_unique[inside_window, ]

# Convert the unique dataframe back to a spatial object
coordinates_bc_albers <- SpatialPoints(coordinates_inside_window, proj4string = CRS(proj4string(coordinates_bc_albers)))
```

```{r}
# Convert the data to a ppp object using the provided Window object
loc_ppp <- ppp(coordinates_bc_albers@coords[,1], coordinates_bc_albers@coords[,2], window = win)

# Plot the ppp object
plot(loc_ppp, pch = 19, cex = 0.5, main = "Spatial Distribution of Moose")
```

According to the ppp plot, we can see a greater concentration of moose moving down towards the south of BC as compared to other regions of BC, suggesting that the density of moose in BC may not be homogenous in the whole province.

## First Moment Descriptive Statistics:

### Intensity

```{r}
# Rescale the window to a specific unit (kilometer from meter)
rescaled_win <- rescale(win, 1000, unitname = "km") 

# Calculate the intensity of mooses per square kilometer
intensity_moose_km2 <- npoints(loc_ppp) / area(rescaled_win)

# Print the result
print(intensity_moose_km2)
```

The estimated intensity of approximately 0.0009469952 moose per square kilometer suggests an extremely low density of moose across British Columbia under the homogenous assumption.

### Quadrant Count

```{r}
# Split the moose locations into a 10 by 10 quadrat and count points
Q <- quadratcount(loc_ppp,
                  nx = 10,
                  ny = 10)

#Quadrat test of homogeneity 
quadrat.test(Q)
```

The small p-value obtained from the quadrat test indicates that the assumption of homogeneity (or complete spatial randomness) is violated.

```{r}
# Plot the moose locations and quadrats
plot(loc_ppp,
     pch = 16,
     cex = 0.5,
     cols = "#046C9A",
     main = "Spatial Distribution of Moose")
plot(Q, cex = 1, col = "red", add = TRUE)
```

```{r}
# Plot the quadrat intensity
plot(intensity(Q, image = TRUE),
     main = "Quadrat Intensity of Moose")

# Plot the moose locations
plot(loc_ppp,
     pch = 16,
     cex = 0.6,
     col = "white",
     add = TRUE)

plot(loc_ppp,
     pch = 16,
     cex = 0.5,
     col = "black",
     add = TRUE)
```

The overall estimated intensity of 0.0009469952 per square kilometer suggests negligible moose presence, which contradicts our knowledge. This discrepancy arises from the flawed assumption of ho- mogeneity in moose distribution, skewing the calculated intensity. The quadrat test reveals a significant p-value, indicating non-homogeneous distribution. Visualizing the data shows varying point counts per quadrat, indicating differing intensities across the area.

### Kernel Density Estimate

```{r}
# Compute the kernel density estimate
density_map <- density.ppp(loc_ppp)

# Plot the kernel density estimate
plot(density_map, main = "Kernel Density Estimate of Moose Spatial Distribution")

# Plot the moose locations
plot(loc_ppp,
     pch = 16,
     cex = 0.6,
     col = "white",
     add = TRUE)

plot(loc_ppp,
     pch = 16,
     cex = 0.5,
     col = "black",
     add = TRUE)
```

From the kernel estimation above, we can see that our data is inhomogeneous, with higher intensity (lambda(u)) in the lower side of the observation window.

```{r}
# Estimate bandwidth using likelihood cross-validation
cv_bandwidth <- bw.diggle(loc_ppp)

# Plot the intensity estimate
plot(density(loc_ppp, sigma = cv_bandwidth), main = "Cross-Validation Bandwidth Selection")
```

```{r}
plot(density(loc_ppp, sigma = bw.ppl), ribbon = F, main = "Likelihood Bandwidth Selection")
```


```{r}
# Estimate R
R <- bw.ppl(loc_ppp)

# Calculate test statistic
LR <- scanLRTS(loc_ppp, r = R)

# Plot the output 
plot(LR, main = "Hotspot Analysis of Moose in BC")
```

```{r}
# Compute local p-values
pvals <- eval.im(pchisq(LR, df = 1, lower.tail = FALSE))

# Plot the output
plot(pvals, main = "Local p-values")
```

```{r}
# Estimate rho
rho <- rhohat(loc_ppp, Elevation)

# Plot the estimated rho function with x-axis limits
plot(rho, main = "Estimated Spatial Dependence between Moose and Elevation", 
     xlim = c(0, max(Elevation)))
```

```{r}
# Estimate rho
rho_f <- rhohat(loc_ppp, Forest)

# Plot the estimated rho function with x-axis limits
plot(rho_f, main = "Estimated Spatial Dependence between Moose and Forest Cover", 
     xlim = c(0, max(Forest)))
```

```{r}
# Estimate rho
rho_f <- rhohat(loc_ppp, HFI)

# Plot the estimated rho function with x-axis limits
plot(rho_f, main = "Estimated Spatial Dependence between Moose and HFI", 
     xlim = c(0, max(HFI)))
```

```{r}
# Estimate rho
rho_d <- rhohat(loc_ppp, Dist_Water)

# Plot the estimated rho function with x-axis limits
plot(rho_d, main = "Estimated Spatial Dependence between Moose and Distance to Water Sources", 
     xlim = c(0, max(Dist_Water)))
```

## Second Moment Statistics

#### Ripley's K-function (Corrected for inhomogeneity)

```{r}
# Estimate a strictly positive density
lambda_ppp_pos <- density(loc_ppp,
                          sigma = bw.ppl,
                          positive = TRUE)

# Simulation envelope (with points drawn from the estimated intensity)
E_ppp_inhom <- envelope(loc_ppp,
                        Kinhom,
                        simulate = expression(rpoispp(lambda_ppp_pos)),
                        correction = "border",
                        nsim = 19,  # alpha level =0.05
                        fix.n = TRUE)

plot(E_ppp_inhom, main = "Inhomogeneous Ripley's K-function with Simulation Envelope", lwd = 2)
```

#### g-function

```{r}
# Estimate a strictly positive density
lambda_ppp_pos <- density(loc_ppp,
                          sigma = bw.ppl,
                          positive = TRUE)

# Simulation envelope for the homogeneous PCF
pcf_homogeneous <- envelope(loc_ppp,
                            pcf,
                            rank = 1,
                            nsim = 19)  # alpha level = 0.05

# Simulation envelope for the inhomogeneous PCF
pcf_inhomogeneous <- envelope(loc_ppp,
                              pcfinhom,
                              simulate = expression(rpoispp(lambda_ppp_pos)),
                              rank = 1,
                              nsim = 19)  # alpha level = 0.05

# Plot the results
plot(pcf_homogeneous, main = "Homogeneous PCF")
plot(pcf_inhomogeneous, main = "Inhomogeneous PCF")
```

```{r}
# Correlation between covariates
cor.im(Elevation, HFI, Dist_Water, Forest, use='na.or.complete')
```

```{r}
# Fit the PPP model
fit <- ppm(loc_ppp ~ Elevation + Dist_Water + Forest, data = DATA)

# Print the fitted model
fit
```

```{r}
# Fit the complex PPP model
fit_complex <- ppm(loc_ppp ~ Elevation + I(Elevation^2) + Dist_Water + Forest + I(Forest^0.5), data = DATA)

# Print the complex model
fit_complex
```

```{r}
AIC(fit); AIC(fit_complex)
```

```{r}
anova(fit, fit_complex, test="LRT")
```

As the p-value is less than 0.05 we reject the simpler model for the more complex one.

```{r}
Q <- quadrat.test(fit_complex, nx = 4, ny = 2)
Q
```

```{r}
plot(Q)
```

```{r}
res <- residuals(fit_complex)

plot(res,
     cols = "transparent")
```

```{r}
plot(parres(fit_complex, 'Elevation'))
```

```{r}
plot(parres(fit_complex, 'Dist_Water'))
```

```{r}
plot(parres(fit_complex, 'Forest'))
```

```{r}
# Generate predictions
predictions <- predict(fit_complex)

# Plot the predictions
plot(predictions, main = 'Predicted Data')
plot(loc_ppp,
     pch = 16,
     cex = 0.8,
     col = "white",
     add = TRUE)
plot(loc_ppp,
     pch = 16,
     cex = 0.6,
     col = "black",
     add = TRUE)
```

## Adding HFI:

```{r}
# Fit the complex PPP model
fit_complex_2 <- ppm(loc_ppp ~ Elevation + I(Elevation^2) + Forest + HFI + I(HFI^3), data = DATA)

# Print the complex model
fit_complex_2
```

```{r}
AIC(fit_complex); AIC(fit_complex_2)
```

```{r}
Q <- quadrat.test(fit_complex_2, nx = 4, ny = 2)
Q
```

```{r}
plot(parres(fit_complex_2, 'Elevation'))
```

```{r}
plot(parres(fit_complex_2, 'Forest'))
```

```{r}
plot(parres(fit_complex_2, 'HFI'))
```

```{r}
# Lurking variable plot for Distance to Water
suppressWarnings({
lurking(fit_complex_2, Dist_Water, type = 'raw', cumulative = F, envelope = T,
        xlab = 'Distance to Water')
})
```

```{r}
# Generate predictions
predictions <- predict(fit_complex_2)

# Plot the predictions
plot(predictions, main = 'Predicted Data')
plot(loc_ppp,
     pch = 16,
     cex = 0.8,
     col = "white",
     add = TRUE)
plot(loc_ppp,
     pch = 16,
     cex = 0.6,
     col = "black",
     add = TRUE)
```
